---
title: "Part-time Workers in the UK"
subtitle: 'Trends from 2018 to 2022'
author: "TimeWise"
date: '`r Sys.Date()`'
output: 
  word_document:
      reference_docx: "Report_template.docx"
      toc: FALSE
      toc_depth: 3

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
	warning = FALSE,
  collapse = TRUE)
```

```{r libraries, include=FALSE}

#### packages needed
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("girdExtra", repos = "http://cran.us.r-project.org")
if(!require(Hmisc)) install.packages("Hmisc", repos = "http://cran.us.r-project.org")
if(!require(sjlabelled)) install.packages("sjlabelled", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(ggtext)) install.packages("ggtext", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
if(!require(doBy)) install.packages("doBy", repos = "http://cran.us.r-project.org")
if(!require(sjmisc)) install.packages("sjmisc", repos = "http://cran.us.r-project.org")
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(flextable)) install.packages("flextable", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
if(!require(ggalt)) install.packages("ggalt", repos = "http://cran.us.r-project.org")
if(!require(janitor)) install.packages("janitor", repos = "http://cran.us.r-project.org")
if(!require(bookdown)) install.packages("bookdown", repos = "http://cran.us.r-project.org")
if(!require(rmdformats)) install.packages("rmdformats", repos = "http://cran.us.r-project.org")
if(!require(htmlwidgets)) install.packages("htmlwidgets", repos = "http://cran.us.r-project.org")
if(!require(widgetframe)) install.packages("widgetframe", repos = "http://cran.us.r-project.org")
if(!require(shiny)) install.packages("shiny", repos = "http://cran.us.r-project.org")

library(ggalt) # contingency table
library(tidyverse) # several tools for data manipulation
library(haven) # read SPSS, Stata, and SAS data
library(gridExtra) # arrange multiple grid-based plots on a page
library(Hmisc) # many functions useful for data analysis
library(sjmisc) # useful to set Na's
library(sjlabelled) # get_labels function
library(ggthemes) # themes for ggplot2
library(ggtext) # fancy text
library(ggrepel) # help with the position of the labels on ggplot2 graph
library(doBy) # 'do' something to data stratified 'by' some variables
library(plotly) # interactive plots
library(flextable) # formatting tables 
library(scales) # add scale symbols and colour to axis
library(janitor) # add totals to tables
library(rmdformats)
library(htmlwidgets)
library(shiny)
library(widgetframe)
```

```{r load data, include=FALSE}

load("../data/LFS_clean.rda")

```



## Background

<p align="justify">People in employment can work on a full-time or a part-time basis. A part-time worker is someone who works fewer hours than a full-time worker. There is no specific number of hours that makes someone full or part-time, but a full-time worker will usually work 35 hours or more a week.</p> 

<p align="justify">This report aims at mapping employment trends for part-time workers in the UK. We use the UK’s largest study on employment circumstances, the Labour Force Survey (LFS), drawing on analysis from the 2018 to 2022 releases (calendar quarters). In the UK Labour Force Survey, the split between full-time and part-time employment is based on respondents’ self-classification. The average hours worked by part-timers, for all the periods included in this analysis, is hours.</p> 

<p align="justify">Our analyses consider only those in employment (current jobs) between 18-64 years old. Table 1 summarises the sample size and population estimates for all the data waves considered in this analysis.</p> 

```{r sample, echo=FALSE}

N_total_2 <- LFS_clean %>%
  filter(AGE >= 18 & AGE <= 64) %>% 
  filter(ILODEFR == 1) %>% # only people in employment
  filter(STAT <= 2) %>% # only employees and self-employed
  filter(STATR <= 2) %>% # only employees and self-employed current job
  group_by(as_label(quarter)) %>%
  summarise(Sample = n(), Population = sum(PWT)) 

N_total_2 %>% flextable() %>% theme_vanilla() %>%
  set_caption(caption = "Table 1: Sample size and population estimates", style = "Table Caption") %>%
  bold(bold = TRUE, part = "header") %>%
  set_header_labels(`as_label(QUARTER)` = "Quarter",
                    Population = "Population estimates")  %>%
  autofit() %>%
  align_nottext_col(align = "center", header = TRUE, footer = FALSE) %>%
  add_footer_lines("Source: UK Labour Force Survey (Person)") %>%
  color(part = "footer", color = "#666666")  

```



## Who is in part-time work?

### By gender

```{r gender, echo=FALSE}



```


### By age

```{r age, echo=FALSE}



```


### By ethnicity 


```{r ethnicity, echo=FALSE}



```


### By qualifications

```{r qualifications, echo=FALSE}



```

### By occupational group

```{r occupations, echo=FALSE}



```


### By intersections 

```{r intersections, echo=FALSE}



```



## What are the reasons for part-time work?

<p align="justify">Part-time jobs are lower paid than full-time jobs, on the whole, and they bring a range of career penalties. There are various reasons for working on a part-time basis. The Labour Force Survey disaggregates part-time employment by reason as follows:</p>

- those who could not find full-time jobs
- those who did not want full-time jobs
- those who were ill or were disabled
- those who were students or were at school

<p align="justify">It also includes those who do not give a reason for part-time working. We focus on those who could not find and did not want full-time jobs in our analysis (involuntary part-time work).</p>


### By sex 

```{r reasons by sex, echo=FALSE}



```

### By age

```{r reasons by age, echo=FALSE}



```

### By ethnicity

```{r reasons by ethnicity, echo=FALSE}



```


### By qualifications

```{r reasons by qualifications, echo=FALSE}



```


### By occupational group

```{r reasons by occupations, echo=FALSE}



```


### By intersections

```{r reasons by intersections, echo=FALSE}



```


## How much do they earn?




## How many hours do they work?


